Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.9)

### 2015-09-29 MSstats_manual_RMarkdown.Rmd
Sections 1-2

The third step, *statistical experimental design*, views the dataset being analyzed as a pilot study of a future experiment, utilizes the variance components of the current datasets, and calculates the minimal number of replicates necessary in the future experiment to achieve a pre-specified statistical power.

@@

`MSstats` takes as input data in a tabular .csv format, which can be generated by any spectral processing tool such as SuperHirn[@Mueller:2007fo], MaxQuant[@Cox:2008ir], Progenesis, MultiQuant, OpenMS[@Sturm:2008eu] or OpenSWATH.

For statistics experts, `MSstats` 3.0 and above satisfies the interoperability requirements of Bioconductor, and takes as input data in the `MSnSet` format[@Gatto:2012tj]. The command line-based workflow is partitioned into a series of independent steps, that facilitate the development and testing of alternative statistical approaches. It complies with the maintenance and documentation requirements of Bioconductor.

@@

`MSstats` is available under the Artistic-2.0 license at [msstats.org](msstats.org). `MSstats` as an external tool is available at [http://proteome.gs.washington.edu/software/Skyline/tools.html](http://proteome.gs.washington.edu/software/Skyline/tools.html). `MSstats` is now also available in [Bioconductor](http:\\www.bioconductor.org). The most recent version of the package is available at [msstats.org](msstats.org) or [MSstats github](https://github.com/MeenaChoi/MSstats). We suggest to use that if possible. The versioning of the main package is updated several times a year, to synchronise with the Bioconductor release.

@@

To help troubleshoot potential problems with installation or functionalities of `MSstats`, a progress report is generated in a separate log file *msstats.log*. The file includes information on the R session (R version, loaded software libraries), options selected by the user, checks of successful completion of intermediate analysis steps, and warning messages. If the analysis produces an error, the file contains suggestions for possible reasons for the errors. If a file with this name already exists in working directory, a suffix with a number will be appended to the file name. In this way a record of all the analyses is kept. Please see the file `KnownIssues-Skyline-MSstatsV2.1.6.pdf` on the `Installation` page of `msstats.org` for a list of known issues and possible solutions for installation problem of MSstats external tool in Skyline

@@

`MSstats` performs statistical analysis steps, that follow peak identification and quantitation. Therefore, input to MSstats is the output of other software tools (such as Skyline or MultiQuant) that read raw spectral files and identify and quantify spectral peaks. The preferred structure of data for use in `MSstats` is  a .csv file in a `long` format with 10 columns. The file should contain the following variables: `ProteinName`, `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, `ProductCharge`, `IsotopeLabelType`, `Condition`, ` BioReplicate`, `Run`, `Intensity`.  The variable names are fixed, but are not case-sensitive. This required input data is generated automatically when using `MSstats` report format in Skyline.

@@

(f) `IsotopeLabelType`: This column indicates whether this measurement is based on the endogenous peptides (use `L`) or labeled reference peptides (use `H`).

(g) `Condition`: For group comparison experiments, this column indicates groups of interest (such as `Disease` or `Control`). For time-course experiments, this column indicates time points (such as `T1`, `T2`, etc). If the experimental design contains both distinct groups of subjects and multiple time points per subject, this column should indicate a combination of these values (such as `Disease_T1`, `Disease_T2`, `Control_T1`, `Control_T2`, etc).

@@

(i) `Run`: This column contains the identifier of a mass spectrometry run. Each mass spectrometry run should have a unique identifier, regardless of the origin of the biological sample. In SRM experiments, if all the transitions of a  biological or a technical replicate are split into multiple `methods` due to the technical limitations, each method should have a separate identifier. When processed by Skyline, distinct values of runs correspond to distinct input file names. It is possible to use the actual input file names as values in the column `Run`.

@@

An example of an input dataset is shown. Example dataset rom an SRM experiment with stable isotope labeled reference peptides. The dataset is stored in a .csv file in a `long` format.  Each row corresponds to a single intensity. More details on assigning the values of `Condition, BioReplicate` and `Run`, depending on the structure of the experimental design, are given below.

@@

In a group comparison design, the conditions (e.g., disease states) are profiled across non-overlapping sets of biological replicates (i.e. subjects). In this example there are 2 conditions, Disease and Control (in general the number of conditions can vary). There are 3 subjects (i.e. biological replicates) per condition (in general an equal number of replicates per condition is not required). Each subject has 2 technical replicate runs (in general technical replicates are not required, and their number per sample may vary). Overall, in this example there are $2 \times 3 \times 2 = 12$ mass spectrometry runs.

@@

The important feature of a time course experimental design is that a same subject (i. e. biological replicate) is repetitively measured across multiple time points. In this example there are 2 time points, Time1 and Time2 (in general the number of times can vary). There are 4 subjects (i.e. biological replicates) measured across times (in general an equal number of times per replicate is not required). There are no technical replicates (in general the number of technical replicates per sample may vary). Overall, in this example there are $2  \times 4 \times 1 = 8$ mass spectrometry runs.

@@

Another frequently used experimental design is a *pared design*, where measurements from multiple conditions (such as healthy biopsy and disease biopsy) are taken from a same subject. The statistical model for this experimental design is the same as in the time course experiment, however the values in the columns of the input data may have a different appearence. In this example there are 2 subjects, PatientA and PatientB (in general the number of patients can vary). There are two conditions per subject, BiopsyHealthy and BiopsyTumor (in general the number of conditions per subject can exceed two). In this example there are 3 technical replicates of each type (in this example, the technical replicates are biopsies; in general these can also be replicate sample preparations or replicate mass spectrometry runs). Overall, in this example there are $2  \times 2 \times 3 = 12$ mass spectrometry runs.

@@

If the data are stored in the format `expressionSet`, group labels information is required. If more than one variable is listed in the argument `group`, then a concatenated variable is created based on all of the specified `group` variables. The remaining information (peptide feature ids, biological replicate ids, and abundance) can be extracted from the rows and columns of `featureData` and `phenoData`, or the users can assign them based on their experimental design.

@@

For label-free DDA experiments the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. In DDA experiments spectral features are defined as peptide ions, which are identified and quantified across runs. Since for label-free DDA experiments some of the columns `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, and `ProductCharge` are not relevant, these columns will have a constant fixed value (such as `NA`) across the entire dataset. Furthermore, the column `IsotopeLabelType` will be set to `L` for the entire dataset.

@@

For label-free DIA experiments, the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. The values of the required columns can be extracted from the output of signal processing software such as Skyline or OpenSWATH.  By default, the combination of the values in the columns `PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge` uniquely identifies each spectral feature (i.e. a fragment ion identified and quantified across multiple runs). If the signal processing software does not provide the information on some of these columns but provides a unique feature identifier, it is possible to use this unique identifier instead of one of these columns. Furthermore, the column `IsotopeLabelType` is set to `L` for the entire dataset.

>>>>

The third step, *statistical experimental design*, views the dataset being analyzed as a pilot study of a future experiment, utilizes the variance components of the current dataset, and calculates the minimal number of replicates necessary in the future experiment to achieve a pre-specified statistical power.

@@

`MSstats` takes as input data in a tabular .csv format, which can be generated by any spectral processing tool such as SuperHirn [@Mueller:2007fo], MaxQuant [@Cox:2008ir], Progenesis, MultiQuant, OpenMS [@Sturm:2008eu] or OpenSWATH.

For statistics experts, `MSstats` 3.0 and above satisfies the interoperability requirements of Bioconductor, and takes as input data in the `MSnSet` format [@Gatto:2012tj]. The command line-based workflow is partitioned into a series of independent steps, that facilitate the development and testing of alternative statistical approaches. It complies with the maintenance and documentation requirements of Bioconductor.

@@

`MSstats` is available under the Artistic-2.0 license at [msstats.org](msstats.org). `MSstats` as an external tool for Skyline is available at [http://proteome.gs.washington.edu/software/Skyline/tools.html](http://proteome.gs.washington.edu/software/Skyline/tools.html). `MSstats` is now also available in [Bioconductor](http://www.bioconductor.org). The most recent version of the package is available at [msstats.org](msstats.org) or [MSstats GitHub](https://github.com/MeenaChoi/MSstats). We suggest to use that if possible. The versioning of the main package is updated several times a year, to synchronise with the Bioconductor release.

@@

To help troubleshoot potential problems with installation or functionalities of `MSstats`, a progress report is generated in a separate log file *msstats.log*. The file includes information on the R session (R version, loaded software libraries), options selected by the user, checks of successful completion of intermediate analysis steps, and warning messages. If the analysis produces an error, the file contains suggestions for possible reasons for the errors. If a file with this name already exists in working directory, a suffix with a number will be appended to the file name. In this way a record of all the analyses is kept. Please see the file `KnownIssues-Skyline-MSstatsV2.1.6.pdf` on the "Installation" page of `msstats.org` for a list of known issues and possible solutions for installation problem of MSstats external tool in Skyline

@@

`MSstats` performs statistical analysis steps, that follow peak identification and quantitation. Therefore, input to MSstats is the output of other software tools (such as Skyline or MultiQuant) that read raw spectral files and identify and quantify spectral peaks. The preferred structure of data for use in `MSstats` is  a .csv file in a "long" format with 10 columns representing the following variables: `ProteinName`, `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, `ProductCharge`, `IsotopeLabelType`, `Condition`, ` BioReplicate`, `Run`, `Intensity`.  The variable names are fixed, but are case-insensitive. This required input data is generated automatically when using `MSstats` report format in Skyline.

@@

(f) `IsotopeLabelType`: This column indicates whether this measurement is based on the endogenous peptides (use "L") or labeled reference peptides (use "H").

(g) `Condition`: For group comparison experiments, this column indicates groups of interest (such as "Disease" or "Control"). For time-course experiments, this column indicates time points (such as "T1", "T2", etc). If the experimental design contains both distinct groups of subjects and multiple time points per subject, this column should indicate a combination of these values (such as "Disease_T1", "Disease_T2", "Control_T1", "Control_T2", etc.).

@@

(i) `Run`: This column contains the identifier of a mass spectrometry run. Each mass spectrometry run should have a unique identifier, regardless of the origin of the biological sample. In SRM experiments, if all the transitions of a  biological or a technical replicate are split into multiple "methods" due to the technical limitations, each method should have a separate identifier. When processed by Skyline, distinct values of runs correspond to distinct input file names. It is possible to use the actual input file names as values in the column `Run`.

@@

An example of an acceptable input dataset is shown below. This example dataset is from an SRM experiment with stable isotope labeled reference peptides. The dataset is stored in a .csv file in a "long" format.  Each row corresponds to a single intensity. More details on assigning the values of `Condition, BioReplicate` and `Run`, depending on the structure of the experimental design, are given below.

@@

In a group comparison design, the conditions (e.g., disease states) are profiled across non-overlapping sets of biological replicates (i.e., subjects). In this example there are 2 conditions, Disease and Control (in general the number of conditions can vary). There are 3 subjects (i.e., biological replicates) per condition (in general an equal number of replicates per condition is not required). Each subject has 2 technical replicate runs (in general technical replicates are not required, and their number per sample may vary). Overall, in this example there are $2 \times 3 \times 2 = 12$ mass spectrometry runs.

@@

The important feature of a time course experimental design is that a same subject (i.e., biological replicate) is repetitively measured across multiple time points. In this example there are 2 time points, Time1 and Time2 (in general the number of times can vary). There are 4 subjects (i.e., biological replicates) measured across times (in general an equal number of times per replicate is not required). There are no technical replicates (in general the number of technical replicates per sample may vary). Overall, in this example there are $2  \times 4 \times 1 = 8$ mass spectrometry runs.

@@

Another frequently used experimental design is a *paired design*, where measurements from multiple conditions (such as healthy biopsy and disease biopsy) are taken from a same subject. The statistical model for this experimental design is the same as in the time course experiment, however the values in the columns of the input data may have a different appearence. In this example there are 2 subjects, PatientA and PatientB (in general the number of patients can vary). There are two conditions per subject, BiopsyHealthy and BiopsyTumor (in general the number of conditions per subject can exceed two). In this example there are 3 technical replicates of each type (in this example, the technical replicates are biopsies; in general these can also be replicate sample preparations or replicate mass spectrometry runs). Overall, in this example there are $2  \times 2 \times 3 = 12$ mass spectrometry runs.

@@

If the data are stored in the format `expressionSet`, information for group labels is required. If more than one variable is listed in the argument `group`, then a concatenated variable is created based on all of the specified `group` variables. The remaining information (peptide feature ids, biological replicate ids, and abundance) can be extracted from the rows and columns of `featureData` and `phenoData`, or assigned by the users based on the experimental design.

@@

For label-free DDA experiments the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. In DDA experiments spectral features are defined as peptide ions, which are identified and quantified across runs. Since for label-free DDA experiments some of the columns `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, and `ProductCharge` are not relevant, these columns will have a constant fixed value (such as `NA`) across the entire dataset. Furthermore, the column `IsotopeLabelType` will be set to "L" for the entire dataset.

@@

For label-free DIA experiments, the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. The values of the required columns can be extracted from the output of signal processing software such as Skyline or OpenSWATH.  By default, the combination of the values in the columns `PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge` uniquely identifies each spectral feature (i.e., a fragment ion identified and quantified across multiple runs). If the signal processing software does not provide the information on some of these columns but provides a unique feature identifier, it is possible to use this unique identifier instead of one of these columns. Furthermore, the column `IsotopeLabelType` is set to "L" for the entire dataset.
